## Predicting Manner for Exercise Activity

Based upon data collected by a group of exercise enthusiasts, the manner in which they preform an exercise activity will be predicted.  Each of 6 participants collected data based on belt, arm, forearm, and dumbbell accelerometers during particular barbell exercises. This data will constitute the training data for the predictive models.

The training data for this analysis is here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

```{r echo=TRUE, eval=TRUE}
library(caret)
```

```{r echo=TRUE, eval=TRUE}
# Load the training dataset
traindf <- read.csv("train/pml-training.csv", header = TRUE, na.strings=c("NA", ""), stringsAsFactors = TRUE)

# Tidy the training data by removing columns with mostly NAs
mostly_good <- apply(!is.na(traindf),2,sum)>=nrow(traindf)
training <- traindf[, mostly_good]

# Remove junk columns including timestamps and non-revelant information from training data
training <- training[, -grep("X|time|window|user_name", names(training))]
```

```{r echo=TRUE, eval=TRUE}
# Load the test dataset
testdf <- read.csv("test/pml-testing.csv", header = TRUE, na.strings=c("NA", ""), stringsAsFactors = TRUE)

# Tidy the testing data by removing columns with mostly NAs
mostly_good <- apply(!is.na(testdf),2,sum)>=nrow(testdf)
testing <- testdf[, mostly_good]

# Remove junk columns including timestamps and non-revelant information from testing data
testing <- testing[, -grep("X|time|window|user_name", names(testing))]
```

```{r echo=TRUE, eval=TRUE}
# Subset the training data to create additional training and testing data sets
set.seed(9876)
splitIndex <- createDataPartition(y=training$classe, p = .001, list = FALSE, times = 1)
cv_training <- training[splitIndex,]
cv_testing <- training[-splitIndex,]
```
The prediction model will be built using the Random Forest method with k fold cross validation on the training data set. This validation method will randomly partition the training set into k equal sized subsamples. For this prediction exercise 5 equal sized training subsamples will be used. The k results from the folds will be averaged to produce a single unbiased estimation of accuracy.
```{r echo=TRUE, eval=TRUE}
# Use Random Forest as the training method to create the prediction model on %75 subset of the training data
train.control <- trainControl(method = "repeatedcv", number = 5)
mf <- train(classe~.,data=cv_training, method="rf", trControl = train.control)
# Plot the predictors in order or relevance
plot(varImp(mf), main = "Variables of Importance")
# Plot the model error
plot(mf, main = "Training Model Accuracy")
```

```{r echo=TRUE, eval=TRUE}
# Predict outcomes based on the testing data set
predictions <- predict(mf, newdata=cv_testing)
print(mf)
```
The out-of-sample error rate will now be calculated using the test data set extracted from the training data set. An unbiased estimate of the Random Forest prediction accuracy is generated.  It is expected that the out-of-sample error rate will be greater than the in-sample-error rate.
```{r echo=FALSE, eval=TRUE}
# Calculate the Out of Sample Error Rate
outofsample.error <- 1 - (sum(predictions == cv_testing$classe) / length(predictions))
paste0("Out of Sample Error: ", round(outofsample.error * 100, digits = 4), " %")
```
